---
title: "Grouping & Summarizing"
author: "STA 1020: Probbility and Statistics I"
output:
  html_document:
    highlight: tango
    theme: united
    toc: no
    toc_depth: 4
    keep_md: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
  md_document:
    toc: yes
    toc_depth: 4
---

Sometimes, one may be interested in aggregations of the data within a given label e.g. each year

For such scenarios one would use the group-by and summarize verbs, which collapse large datasets into manageable summaries.

Besides the `mean()` and `sum()` functions for summarizing a collection of numbers, there is also the `median()` in combination with `summarize()`

Use the `median()` function within the `summarize()` to find the median GDP per capita. Save it into a column called `mediangdpPercap`

```{r, results="show", message=FALSE, warning=FALSE}
library(gapminder)
library(dplyr)

gapminder %>%
  summarize(mediangdpPercap = median(gdpPercap))
```

### Summarizing the median GDP per capita in 2007

One can also sub-set the data to obtain the median GDP per capita for a given time/ year

1)  Obtain the subset of the data for the year 2007

2)  use the `median()` function within the `summarize()` to obtain the median GDP per capita

3)  save the output into the column `mediangdpPercap`

```{r, results="show", message=FALSE, warning=FALSE}
library(gapminder)
library(dplyr)

gapminder %>%
  filter(year == 2007) %>%
  summarize(mediangdpPercap = median(gdpPercap))
```

### Summarizing multiple variables in 2007

The `summarize()` verb allows you to summarize multiple variables at once.

In this case, you'll use the `median()` function to find the median life expectancy and the `max()` function to find the maximum GDP per capita.

1)  find both the median life expectancy (*`lifeExp`*) and the maximum GDP per capita (*`gdpPercap`*) in the year 2007

2)  assign them to `medianLifeExp` and `maxGdpPercap` respectively

3)  you can use the `max()` function to find the maximum

```{r, results="show", message=FALSE, warning=FALSE}
library(gapminder)
library(dplyr)

gapminder %>%
  filter(year == 2007) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))
```

### Summarizing by year

We have obtained the median life expectancy and the maximum GDP per capita in the year 2007

1)  perform a similar summary within each year in the dataset using the `group-by` verb

2)  assign the output to `medianLifeExp` *and* `maxGdpPercap`, respectively.

```{r, results="show", message=FALSE, warning=FALSE}
library(gapminder)
library(dplyr)

gapminder %>%
  group_by(year) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))
```

Now compare the above output among continents: subset by the year 2007.

```{r, results="show", message=FALSE, warning=FALSE}
library(gapminder)
library(dplyr)

gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))
```

Instead of grouping just by year, or just by continent, group by both continent and year to summarize within each.

Assign the outcome to medianLifeExp`and`maxGdpPercap\`, respectively

```{r, results="show", message=FALSE, warning=FALSE}
library(gapminder)
library(dplyr)

gapminder %>%
  group_by(continent, year) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))
```

### Visualizing median life expectancy over time

In the last chapter, you summarized the gapminder data to calculate the median life expectancy within each year. This code is provided for you, and is saved (with `<-`) as the `by_year` dataset.

Now you can use the `ggplot2` package to turn this into a visualization of changing life expectancy over time.

1)  use the `by_year` dataset to create a scatter plot showing the change of median life expectancy over time

2)  put the `year` on the x-axis and medianLifeExp on the y-axis

3)  be sure to add `expand_limits(y = 0)` to make sure the plot's y-axis includes zero.

```{r, results="show", message=FALSE, warning=FALSE}
library(gapminder)
library(dplyr)
library(ggplot2)

by_year <- gapminder %>%
  group_by(year) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))
by_year

# Create a scatter plot showing the change in medianLifeExp over time
ggplot(by_year, aes(x = year, y = medianLifeExp)) + 
  geom_point() + 
  expand_limits(y = 0)
```

Summarize the gapminder dataset by continent and year, finding the median GDP per capita (`gdpPercap`) within each and putting it into a column called `medianGdpPercap`

1)  Use the assignment operator `<-` to save this summarized data as `by_year_continent`

2)  Create a scatter plot showing the change in `medianGdpPercap` by continent over time

3)  Use color to distinguish between continents, and remember to add `expand_limits(y = 0)` so that the y-axis starts at zero.

```{r, results="show", message=FALSE, warning=FALSE}
library(gapminder)
library(dplyr)
library(ggplot2)

by_year_continent <- gapminder %>%
  group_by(continent, year) %>%
  summarize(medianGdpPercap = median(gdpPercap))

ggplot(by_year_continent, aes(x = year, y = medianGdpPercap, color = continent)) + 
  geom_point() + 
  expand_limits(y = 0)
```

### Comparing median life expectancy and median GDP per continent in 2007

1)  filter the gapminder dataset for the year 2007

2)  summarize the median GDP per capita and the median life expectancy within each continent, into columns called `medianLifeExp` and `medianGdpPercap`

3)  Save this as `by_continent_2007`

4)  Use the `by_continent_2007` data to create a scatterplot comparing these summary statistics for continents

5)  put the median GDP per capita on the x-axis and the median life expectancy on the y-axis

6)  Color the scatter plot by `continent`

```{r, results="show", message=FALSE, warning=FALSE}
library(gapminder)
library(dplyr)
library(ggplot2)

# Summarize the median GDP and median life expectancy per continent in 2007
by_continent_2007 <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(medianLifeExp = median(lifeExp),
            medianGdpPercap = median(gdpPercap))

# Use a scatter plot to compare the median GDP and median life expectancy
ggplot(by_continent_2007, aes(x = medianGdpPercap, y = medianLifeExp, color = continent)) + 
  geom_point()
```
